---
title: "Getting Started with RankMap"
author:
  - name: Jinming Cheng
    affiliation:
    - &nus_cbds Centre for Biomedical Data Science,
                Duke-NUS Medical School,
                Singapore 169857,
                Singapore
               
    # email: jinming.cheng@outlook.com
    

date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Getting Started with RankMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

output:
  BiocStyle::html_document:
    toc_float: false
bibliography-style: plain
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    eval = TRUE, 
    collapse = TRUE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 60,
    fig.height = 4.2,
    fig.width = 5.6
)
```


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Introduction

**RankMap** is an R package for fast, robust, and scalable reference-based 
cell type annotation in single-cell and spatial transcriptomics data. 
It works by transforming gene expression matrices into sparse ranked 
representations and training a multinomial logistic regression model 
using the `glmnet` framework. This rank-based approach improves 
robustness to batch effects, platform differences, and partial gene 
coverage—especially beneficial for technologies such as Xenium and MERFISH.

**RankMap** supports commonly used data structures 
including `Seurat`, `SingleCellExperiment`, and `SpatialExperiment`. 
The workflow includes flexible preprocessing steps such as 
top-K gene masking, binning, expression weighting, and scaling, 
followed by efficient model training and rapid prediction.

Compared to existing tools such as **SingleR**, **RCTD** (via **spacexr**),
and **Azimuth**, **RankMap** achieves comparable or superior accuracy 
with significantly faster runtime, making it particularly well suited 
for high-throughput applications on large datasets.

This vignette provides a quick-start guide to using **RankMap** for 
cell type prediction.


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Installation

Install **RankMap** from GitHub:

```{r, eval=FALSE}
devtools::install_github("jinming-cheng/RankMap")
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Quick Start (Seurat Objects)

## Load Data

```{r}
library(RankMap)
```

Load example single-cell RNA-seq dataset (17,597 genes x 150 cells):

```{r}
seu_sc <- readRDS(system.file("extdata", "seu_sc.rds", package = "RankMap"))
seu_sc
```

Load example Xenium spatial transcriptomics dataset (313 genes x 150 cells):

```{r}
seu_xen <- readRDS(system.file("extdata", "seu_xen.rds", package = "RankMap"))
seu_xen
```

## Predict Cell Types

Run cell type prediction using the `RankMap()` function. 
By default, RankMap uses normalized expression from the "data" slot. 
For spatial datasets with limited gene panels, 
a smaller `k` (e.g., `k = 20`) is typically sufficient. 
For single-cell RNA-seq with deeper coverage, 
larger values of `k` (e.g., 100 or 200) are generally recommended.

```{r}
pred_df <- RankMap(
  ref_data = seu_sc,
  ref_labels = seu_sc$cell_type,
  new_data = seu_xen,
  k = 20
)
```

The result is a `data.frame` containing: 
`cell_id`, `predicted_cell_type` and `confidence`

```{r}
head(pred_df)
```

## Evaluate Performance

If ground truth labels are available, you can evaluate prediction accuracy using:

```{r}
perf <- EvaluatePredictionPerformance(
  prediction_df = pred_df,
  truth = seu_xen$cell_type_SingleR
)
perf
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->



<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

# Quick Start (SummarizedExperiment Objects)

## Prepare Data

Convert `Seurat` objects into `SingleCellExperiment` objects:

```{r}
library(SingleCellExperiment)

sce_sc <- SingleCellExperiment(
  assays = list(
    counts = GetAssayData(seu_sc, layer = "counts"),
    logcounts = GetAssayData(seu_sc, layer = "data")
  ),
  colData = seu_sc@meta.data
)

sce_sp <- SingleCellExperiment(
  assays = list(
    counts = GetAssayData(seu_xen, layer = "counts"),
    logcounts = GetAssayData(seu_xen, layer = "data")
  ),
  colData = seu_xen@meta.data
)
```

## Predict Cell Types

Run cell type prediction using the `RankMap()` function. 
Set `k = 100` as a reasonable default when the optimal number of top-ranked genes is unknown. 
When using `SummarizedExperiment` input, the `logcounts` assay is used automatically.

```{r}
pred_df <- RankMap(
  ref_data = sce_sc,
  ref_labels = sce_sc$cell_type,
  new_data = sce_sp,
  k = 100
)
```

## Evaluate Performance

Compare predictions with ground truth labels:

```{r}
perf <- EvaluatePredictionPerformance(
  prediction_df = pred_df,
  truth = sce_sp$cell_type_SingleR
)
perf
```


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


# Session Info

```{r}
sessionInfo()
```

\pagebreak
